---
title: "Econ 97 Labs"
author: "Patrick Molligo"
#date: "`r Sys.Date()`"
date: "`r format(Sys.Date(), '%B %d, %Y')`"
output: pdf_document
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE
)
```

# How to Use This Lab Manual

These labs are designed to reinforce *probability theory* by using **simulation** and **computational verification** in R.

Why simulation helps in probability:

* Many probabilities are defined as *long-run frequencies* (e.g., the chance of heads is the limiting proportion of heads in repeated coin flips).
* Simulation lets us **approximate** theoretical probabilities and see concepts like the **Law of Large Numbers** and **Central Limit Theorem** in action.
* Simulation is also a practical tool when exact calculations are difficult.

Throughout, we will use:

* **Random sampling** (`sample()`, `rbinom()`, etc.)
* **Indicators** like `(event)` which evaluate to `TRUE/FALSE`, and means of indicators to estimate probabilities
* **Repetition** (`replicate()`) to generate many trials
* **Visualization** (`plot()`, `hist()`) to connect distributions to data

---

# Lab 1 — R Basics + Simulation + Law of Large Numbers

## Objectives

* Learn R basics: vectors, arithmetic, logical comparisons, and summaries
* Estimate probabilities through simulation
* See the **Law of Large Numbers (LLN)**: the sample proportion approaches the true probability as sample size grows

## Concepts

### Probability as a long-run frequency

If we repeat an experiment many times, the proportion of times an event occurs tends to stabilize around the true probability. In simulation, we approximate a probability by:


$$P(A) \approx \frac{{\text{trials where }A\text{ happens}}}{\text{number of trials}}.$$


### Indicator variables

If we define an indicator (I_A) that equals 1 when event (A) occurs and 0 otherwise, then:

* The average of (I_A) across many trials estimates (P(A)).
* In R, `(condition)` produces `TRUE/FALSE`, and `mean(TRUE/FALSE)` treats `TRUE` as 1 and `FALSE` as 0.

## Warm-up (run each line)

```{r}
1:10
sum(1:10)
mean(1:10)
c(2, 5, 9)^2
```

## Part A — Coin flips and LLN

Simulate coin flips and estimate (P(\text{Heads})).

```{r}
set.seed(1)
flips <- sample(c("H","T"), size = 50, replace = TRUE)
flips
mean(flips == "H")
```

Repeat with larger sample sizes and observe stabilization.

```{r}
set.seed(1)
for (n in c(50, 500, 5000, 50000)) {
  flips <- sample(c("H","T"), size = n, replace = TRUE)
  cat("n =", n, " prop_heads =", mean(flips == "H"), "\n")
}
```

### Running estimate (visual LLN)

A running estimate shows how the proportion evolves with each additional flip.

```{r}
set.seed(2)
n <- 5000
x <- sample(c(1,0), size = n, replace = TRUE)  # 1=heads, 0=tails
running <- cumsum(x) / (1:n)
plot(running, type="l", xlab="Flip number", ylab="Running proportion of heads")
abline(h=0.5, lty=2)
```

## Part B — Multiple-choice guessing

A 30-question quiz, 4 choices each. If you guess randomly, each question is correct with probability (p = 1/4). The total score is:

$$X \sim \text{Binomial}(n=30, p=1/4).$$

The theoretical expected score is ($E[X] = np = 30 \cdot 1/4 = 7.5$).

Simulate 10,000 students:

```{r}
set.seed(3)
B <- 10000
scores <- replicate(B, sum(sample(1:4, 30, replace=TRUE) == 1))
mean(scores)
```

Estimate ($P(X \ge 12)$):

```{r}
mean(scores >= 12)
```

### Deliverables

* A short description (2–4 sentences) of what you see in the LLN running plot
* Estimated expected score and ($P(\text{score} \ge 12)$)
* Compare the simulated mean to the theoretical mean 7.5

---

# Lab 2 — Counting + Sampling Without Replacement

## Objectives

* Model draws **without replacement**
* Estimate conditional probabilities from sequential experiments
* Connect to **hypergeometric** reasoning and conditional probability rules

## Concepts

### Sampling without replacement

When items are not replaced, outcomes are **dependent**. For example, after drawing a sour milk, there are fewer sour milks remaining.

### Conditional probability

$$P(A \mid B) = \frac{P(A \cap B)}{P(B)}.$$
In simulation, we estimate ($P(A \mid B)$) by restricting to the trials where (B) happened:
$$P(A \mid B) \approx \frac{{\text{trials where }A\text{ and }B}}{{\text{trials where }B}}.$$

## Part A — One draw

Urn contains 7 good (G) and 3 sour (S).

```{r}
urn <- c(rep("G", 7), rep("S", 3))
set.seed(10)
draw <- sample(urn, size=1, replace=FALSE)
draw
```

Simulate many single draws to estimate (P(S)).

```{r}
set.seed(11)
B <- 10000
draws <- replicate(B, sample(urn, 1))
mean(draws == "S")
```

## Part B — Sequential customers (conditional probability)

Simulate 3 sequential purchases (without replacement). Estimate:
$$P(\text{3rd is sour} \mid \text{2nd is not sour}).$$

```{r}
set.seed(12)
B <- 20000

second_not_sour <- logical(B)
third_sour <- logical(B)

for (b in 1:B) {
  purchase <- sample(urn, size=3, replace=FALSE)
  second_not_sour[b] <- (purchase[2] != "S")
  third_sour[b] <- (purchase[3] == "S")
}

mean(third_sour[second_not_sour])
```

### Deliverables

* Estimated ($P(S)$) for the first draw
* Estimated ($P(\text{3rd sour} \mid \text{2nd not sour})$)
* 2–3 sentences on why conditioning changes the probability

---

# Lab 3 — Discrete Random Variables: Binomial + Visualization

## Objectives

* Use `dbinom`, `pbinom`, `rbinom`
* Compare analytic probabilities to simulation
* Visualize a probability mass function (PMF)

## Concepts

### Binomial distribution

If (X) counts successes in (n) independent trials with success probability (p), then:
$$P(X=k) = \binom{n}{k} p^k (1-p)^{n-k}.$$
In R:

* `dbinom(k, n, p)` gives ($P(X=k)$)
* `pbinom(k, n, p)` gives ($P(X \le k)$)
* `rbinom(B, n, p)` simulates (B) values of (X)

## Part A — Exact probabilities

Let ($X \sim \text{Binomial}(n=20, p=0.4)$).

```{r}
dbinom(8, size=20, prob=0.4)          # P(X=8)
pbinom(8, size=20, prob=0.4)          # P(X<=8)
pbinom(10, 20, 0.4) - pbinom(5, 20, 0.4)  # P(6<=X<=10)
```

## Part B — Simulation check

```{r}
set.seed(20)
x <- rbinom(50000, size=20, prob=0.4)

mean(x == 8)
mean(x <= 8)
mean(x >= 6 & x <= 10)
```

## Part C — Plot the PMF

```{r}
k <- 0:20
pmf <- dbinom(k, 20, 0.4)
plot(k, pmf, type="h", xlab="k", ylab="P(X=k)")
points(k, pmf)
```

### Deliverables

* Three probabilities (analytic + simulation)
* One paragraph explaining what a CDF is in words

---

# Lab 4 — Geometric & Poisson + Approximations

## Objectives

* Understand waiting times (geometric)
* Understand counts over intervals (Poisson)
* Compare analytic results to simulation

## Concepts

### Geometric distribution (waiting time)

If each trial succeeds with probability (p), and (T) is the number of trials until first success, then:
$$E[T] = \frac{1}{p}.$$
**R note:** `rgeom()` returns the number of failures before first success, so we add 1 to get trials-until-success.

### Poisson distribution (counts)

If events happen randomly at average rate ($\lambda$) in an interval, then ($Y \sim \text{Poisson}(\lambda)$). In R:

* $dpois(y, lambda)$ gives ($P(Y=y)$)
* $ppois(y, lambda)$ gives ($P(Y \le y)$)
* $rpois(B, lambda)$ simulates values

## Part A — Geometric waiting time

```{r}
set.seed(30)
p <- 0.2
T <- rgeom(20000, prob=p) + 1
mean(T)   # compare to 1/p = 5
mean(T > 8)
```

## Part B — Poisson probabilities

Let ($Y \sim \text{Poisson}(\lambda=3)$). Compute and simulate:

* ($P(Y=0)$)
* ($P(Y \ge 5)$)

```{r}
dpois(0, lambda=3)
1 - ppois(4, lambda=3)

set.seed(31)
Y <- rpois(50000, lambda=3)
mean(Y == 0)
mean(Y >= 5)
```

### Deliverables

* Simulated vs theoretical (E[T])
* Two Poisson probabilities (analytic + simulation)

---

# Lab 5 — Continuous Random Variables + Normal + Central Limit Theorem

## Objectives

* Use `pnorm`, `rnorm`, `runif`
* Compute Normal probabilities
* See the **Central Limit Theorem (CLT)** using sample means

## Concepts

### Normal distribution and tail probabilities

For a standard normal ($Z \sim N(0,1)$):

* `pnorm(a)` computes ($P(Z \le a)$)
* tail probabilities use `1 - pnorm(a)`

### CLT (informal)

For many i.i.d. random variables with mean ($\mu$) and standard deviation ($\sigma$), the sample mean ($\bar X$) becomes approximately Normal for large (n):
$$\bar X \approx N\left(\mu,\ \frac{\sigma}{\sqrt{n}}\right).$$

## Part A — Normal probabilities

```{r}
pnorm(1.25)                  # P(Z<1.25)
2*(1 - pnorm(2))             # P(|Z|>2)
```

## Part B — CLT with dice (distribution of sample means)

```{r}
set.seed(40)
B <- 20000
ns <- c(1, 5, 20, 50)

means_list <- lapply(ns, function(n) {
  replicate(B, mean(sample(1:6, size=n, replace=TRUE)))
})

par(mfrow=c(2,2))
for (i in seq_along(ns)) {
  hist(means_list[[i]], main=paste("n =", ns[i]), xlab="sample mean", breaks=40)
}
par(mfrow=c(1,1))
```

## Part C — Compare simulation to Normal approximation (optional)

For a fair die:

* ($\mu = 3.5$)
* ($\mathrm{Var}(X) = 35/12), so (\sigma = \sqrt{35/12}$)

Compare $P(\bar X > 4)$ for (n=50).

```{r}
n <- 50
xbar <- means_list[[which(ns==n)]]
p_sim <- mean(xbar > 4)

mu <- 3.5
sigma <- sqrt(35/12)
p_norm <- 1 - pnorm(4, mean=mu, sd=sigma/sqrt(n))

p_sim
p_norm
```

### Deliverables

* Two Normal probability answers
* The 2×2 histogram figure for CLT
* Simulation vs Normal approximation for (n=50)

---

# Lab 6 — Estimation + Bootstrap Confidence Intervals

## Objectives

* Understand sampling variability for an estimator (like the mean)
* Use the **bootstrap** to build a confidence interval

## Concepts

### Sampling variability

Even if the population is fixed, different samples produce different sample means.

### Bootstrap idea

We approximate the sampling distribution of an estimator by:

1. Taking your observed sample
2. Resampling it *with replacement* many times (bootstrap samples)
3. Recomputing the estimator (mean) each time

A percentile bootstrap CI uses quantiles of bootstrap estimates.

## Part A — Create a population and take a sample

We use an exponential distribution as a "population" because it is skewed (helpful for seeing why Normal approximations sometimes struggle with small samples).

```{r}
set.seed(50)
pop <- rexp(200000, rate=1)  # true mean is 1
sample1 <- sample(pop, size=50, replace=FALSE)
mean(sample1)
```

## Part B — Bootstrap CI for the mean

```{r}
set.seed(51)
B <- 5000
boot_means <- replicate(B, mean(sample(sample1, size=length(sample1), replace=TRUE)))

ci <- quantile(boot_means, c(0.025, 0.975))
ci

hist(boot_means, breaks=40, main="Bootstrap means", xlab="mean")
abline(v=ci, lty=2)
```

## Part C — Coverage experiment (optional challenge)

Repeat many times to estimate how often a 95% bootstrap CI contains the true mean 1.

```{r}
set.seed(52)
R <- 200     # number of repeated experiments
B <- 2000    # bootstrap reps per experiment

covers <- logical(R)

for (r in 1:R) {
  sample1 <- sample(pop, size=50, replace=FALSE)
  boot_means <- replicate(B, mean(sample(sample1, size=length(sample1), replace=TRUE)))
  ci <- quantile(boot_means, c(0.025, 0.975))
  covers[r] <- (ci[1] <= 1) & (1 <= ci[2])
}

mean(covers)
```

### Deliverables

* Your bootstrap 95% CI
* 3–5 sentences interpreting what the CI means (and what it does *not* mean)

---

# Lab 7 — Hypothesis Testing via Simulation (p-values)

## Objectives

* Interpret p-values as probabilities under the null model
* Compute p-values by exact calculation and by simulation
* Visualize the null distribution

## Concepts

### Null hypothesis and p-value

A hypothesis test begins with a null model ($H_0$). A p-value is:

> The probability, under ($H_0$), of seeing a result at least as extreme as the one observed.

This is a probability *about the data given the model*, not the probability that the model is true.

## Scenario: Is a die biased toward 6?

Roll a die 60 times and observe 16 sixes.

Under ($H_0$): fair die, (p = 1/6), and
$$X \sim \text{Binomial}(n=60, p=1/6).$$

## Part A — Exact one-sided p-value

Compute ($P(X \ge 16)$).

```{r}
1 - pbinom(15, size=60, prob=1/6)
```

## Part B — Simulation p-value

```{r}
set.seed(60)
B <- 200000
X <- rbinom(B, size=60, prob=1/6)
p_sim <- mean(X >= 16)
p_sim
```

## Part C — Visualize the null distribution

```{r}
hist(X, breaks=50, main="Null distribution of # of sixes (fair die)", xlab="# of sixes")
abline(v=16, lty=2)
```

### Deliverables

* Exact p-value and simulation p-value
* 3–5 sentences interpreting the p-value in plain English

---

# General Rubric (apply to all labs)

* Code runs and is organized (30%)
* Numerical answers correct (30%)
* Written interpretation connects to probability concept (30%)
* Presentation: labeled axes, readable output, brief explanations (10%)


